### 2017-07-22 chapters/introduction/patient.tex
Add latest

\section{Personalization of Cardiac Biomechanics}

In this section we will go through the steps needed to personalize the
mechanical models developed in the previous section towards a given
patient. We will start by giving a brief overview of the
interaction between medical imaging and cardiac computational model.
Then we will explain how the geometries are extracted from the
medical images and turned into a computational mesh,  which we can use
for finite element computations. Finally we will explain how to
incorporate additional data into model by means of adjoint based data
assimilation techniques.

@@

Cardiac computational modeling often relies on data obtained using
medical imaging techniques, and the quality of such data is therefore of
fundamental importance \cite{lamata2014images}. It is clear, that
without advancements in medical imaging technology, the role of
computational models would be limited to simple, idealized cases. With
access to high quality images, detailed information about cardiac structure
can be used to e.g build anatomically correct models. The
computational models can then later be validated agains this images,
or we can assimilate additional data extracted from this images to
extract biomarkers. Figure \ref{fig:cardiac_imaging_model} summarizes
this interaction between medical images and computational models.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used togther with the medical images to
  assimilate additional data in order to extraction clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

% Still, one of the major bottlenecks in terms of personalized
% computaional models is the quality of data (and representations of
% model). There is a big issue in
% reproducebility, because of noisy data, and operator depedent
% measurements.
Today there are three main non-invasive imaging techniques used in
cardiology. These are chocardiography (Echo), magnetic
resonance imaging (MRI) and computed tomography (CT). Each
modality offers advantages and disadvatages over the other

For example, MRI provides
high quality images, uses zero radiation, but is exspensive and lacks
temporal resolution. CT can more accuratley reconstruct the 3D image
in contrast to MRI, in which 2D slices needs to be glued together to
form a 3D surface. However, CT exposes the patient to radiation which
increases the chance of developing cancer. Finally echocardiography is
easy to use, cheap, harmless, and  has good temporal resolution, but
is clearly inferior when it comes to image quality.

@@

The main modality used in this thesis is 4D echocardiography, and we
will therefore focus on data aquired using this modality.
With 4D we mean three spatial and one temporal dimesion.
The speed of sound in human tissue (which is approxmately 1540 m/s)
put some limitations on the image quality verus the frame-rate
\cite{rabben2010technical} for these 3D volume images. In order to
increase the resolution, such an image is formed by stiching together
$N$ disjoint subvolumes aquired during $N$ cardiac cycles ($N$
typicall between 2 and 8)\cite{brekke2007volume}.

The images are later processed using some image segmentation tool.
One such tool is EchoPac which is used for analysing images for GE
Vingmed. The 4D Auto LVQ tool in EchoPac, is a tool for processing 3D
echo images, and traces of volume, triangulated surfaces and 3D strain
\cite{heimdal20114d} traces together with a structured mesh of the
American Heart Association (AHA) segments
\cite{cerqueira2002standardized} for each time point, see Figure
\ref{fig:echopac_output}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.38\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/bullseye/bullseye.pdf}
    \caption{\label{fig:bullseye_intro}}
  \end{subfigure}
\caption{Figure \ref{fig:echopac_out_surf} shows triangulated surfaces of the
      endo- and epicardium, and Figure
      \ref{fig:echopac_out_strain_mesh} shows structured mesh of the
      American Heart Association (AHA) segments. These surfaces are
      exported from the image segmentational tool EchoPac. Figure 
      \ref{fig:bullseye_intro} shows the American Heart Association (AHA)
      segments illustrated on a so called bullseye plot with name and
      numbers on the different segments. }
\label{fig:echopac_output}
\end{figure}

@@

\subsubsection{Mesh generation}
In this section we will explain how to generate a left ventricular
mesh based on segmented surfaces coming from 4D echocardiography.
Figure \ref{fig:echopac_output} shows an example of how the exported data
from the image semgentation tool looks like. For each time point we
can extract triangulated surfaces of the endocardium and epicardium
which can be seen in Figure \ref{fig:echopac_out_surf}. Together with
these surfaces we are also given a so called strain mesh (Figure
\ref{fig:echopac_out_strain_mesh}) located approximately in the
midwall, which defines the approximate location of the AHA-segments
\cite{cerqueira2002standardized}, and can be used to orient the mesh.

As discussed in \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the postive $x$
direction. To conststruct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
sqauare fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct size is found the points above the plane is removed,
and the remaining surfaces are smoothed using the GAMer
\cite{yu2008feature}.

@@

The actualy mesh generation is performed using Gmsh
\cite{geuzaine2009gmsh}, which meshes the endocardial and epicardial
surface togther unsing frontal-Delaunay meshing algorithm. Gmsh also marks
the endocardal, the epicardial  and the basal facets, along with the
endocardial and epicardial basal rings.

@@

\begin{remark}
  The cavity volume is computed by generating the whole mesh, and the
  compute the volume using Equation \eqref{eq:volume_operator}. The cut size is
  found by using a one dimensional optimization algorithm with
  the objective funcitonal representing the squared error between
  computed and measured volume.
\end{remark}

\subsubsection{Rule-based fiber architecture}
\label{sec:rule_based_fiber}
Both the passive and active properties of the myocardium depends on
the underlying muscle fiber architecture, and it is therefore of high
relevance to capture this architecture as accuratley as
possible. Unfortunately, imaging technology does not currently provide a
way of extracting patients-specific fiber ortiention \emph{in
  vivo}. Diffusion Tensor Imaging (DTI) can be utilized to reconstruct
the fiber and lamiar structure \cite{rohmer2007reconstruction}
\emph{ex vivo}.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]
  {chapters/introduction/figures/fiber/fiber}
\caption{The LDRB algorithm \cite{bayer2012novel} applied to an LV
  (left) and BiV (right) geometry with an helix fiber angle of
  $60^{\circ}$ and $-60^{\circ}$ on the endo- and epicardium
  respectively. Color represents the helical angle, with blue being
  $0^{\circ}$ (circumferential) which can be seen in the midwall. }
\label{fig:echopac_output}
\end{figure}

The LDRB algorithm takes as
input the geometry $\Omega$, the helical fiber angle on the endo- and
epicardium ($\alpha_{\mathrm{endo}}$ and $\alpha_{\mathrm{epi}}$) and
the transverse fiber angle on the endo- and epicardium
($\beta_{\mathrm{endo}}$ and $\beta_{\mathrm{epi}}$), and outputs a
set of three orthogonal vector fields, one representing the fiber
angle $(\ef)$, one representing the sheet angle $(\es)$ and the final
diraciton is called the sheet normal direction $(\en)$

@@

When dealing with geometric shapes such as a left ventricle, finding a
coordinate system that is easy to work with is important. For
example, when studing spherical shapes, the spherical coordinate system is
usually easier to work with. Likewise, a coordinate system that is
easy to use when studying ellipsoidal shapes like the left ventricle is
the \emph{prolate spheroidal coordinate system} \cite{hunter1996kd}. 
The cartesian coordinates $(x,y,z)$ are related to the prolate
spheroidal coordinates by

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta  \\
    y &= a \sinh \mu \sin \nu \sin \theta  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuthal angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semimajor
axis, and $c$ the semiminor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. There inverse relation, going from the
cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$ is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau) \\
    \mu &= \arccosh (\sigma) \\
    \theta &= \arctan \frac{z}{y} \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big) \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big) 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{}
\label{fig:prolate}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
The choice of reference geometry ..
 choosing the geometry with the smallest
volume as the reference geometry (i.e the end-systolic geometry)
(references), the mid-diastolic or the end-diastolic ??

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood pressure.
This is analogous to finding the geometry of a deflated balloon, given
an image of and inflated one. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to obtain the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some medical image, and
is thus loaded with some loads $\mathbf{T}$, and we want to find the unloaded geometry
$\Omega_{\mathcal{U}}$, which is so that when we load $\Omega_{\mathcal{U}}$ with $\mathbf{T}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{some_force_balance_eq} with $\mathbf{T}$ representing
the traction on the endocardium, and which depends on the reference
coordinates. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To
find  $\Xvec_{\mathcal{U}}$ we let
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the load $\mathbf{T}$ do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge especially if the tissue is soft and the pressure in
the right ventricle is large. In this case it can happen that the
right ventricle free wall collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides an a more stable
way of obtainind an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}. Let $k$ be a scalar, and suppose we can write
\begin{align}
  \Xvec_{\mathcal{I}} = \Xvec_{\mathcal{U}} + k \cdot \mathbf{d}(\Xvec_{\mathcal{U}}).
\end{align}
Of course this assumption might not be true, but we can find a value of
$k$ that minimizes some residual using a 1D optimization
algorithm. Note that choosing $k = 1$, would be the same as applying
one iteration of the backward displacement method.

Since the backward displcement method really does converge towards a
fixed point, it is favorable to the Rahavan method. Therefore, a
hybrid approach is to apply the backward displacement method as
long as the non-linear solver converge, and switch to the Rahavan
method upon divergence.

@@

For a transversely isotropic material (or orthotropic), the change in
fiber field when going from one geometry to another, has to be
considered. The natural thing to do is apply a \emph{Piola
  transformation} to the fibers in the image based geometry: Let
$\F_{n}= \I - \Grad \left(\mathbf{d}(\Xvec_n) \right)$ and left
$\ef^{n} = J^{-1}\F_{n} \ef$ where $\ef$ is the fiber field on
$\Omega_{\mathcal{I}}$. Another way is to regenerate the fibers (see
Section \ref{sec:rule_based_fiber}) using the new unloaded geometry as
input. We will see in the following numerical experiment that non of
these methods gives the desired result.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stresses in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the contitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical system such as the heart or weather
forecast it is not allways possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

@@

There are basically two main approaches to assimilate data;
\emph{sequential data assimilation} and \emph{variational data
  assimilation}. Sequential data assimilation, which is often
referred to as filtering, is based on a predictor-corector approach
where you start from your initial data and use a filter to predict the
next state. Once you have your measurement available you correct the
estimate based on the new observations. Some examples of sequential
procedures include Kalman filtering and Luenberger observers
\cite{chapelle2013fundamental}.

>>>>

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical model developed in the previous section. We will start by
giving a brief overview of the interaction between medical imaging and
cardiac computational modeling. Then we will explain how the geometries
are extracted from the medical images and turned into a computational
meshes which we can use for finite element computations. Finally we
will explain how to incorporate additional data into model by means of
adjoint based data assimilation techniques.

@@

Cardiac computational modeling often relies on data obtained using
medical imaging techniques, and the quality of such data is therefore of
fundamental importance \cite{lamata2014images}. It is clear, that
without advancements in medical imaging technology, the role of
computational models would be limited to simple, idealized cases. With
access to high quality images, detailed information about cardiac structure
can be used to e.g build anatomically correct models. The
computational models can then later be validated using the same
images, or additional data can be extracted from the images in order to
compute potential biomarkers. Figure \ref{fig:cardiac_imaging_model} summarizes
this interaction between medical images and computational models.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used together with the medical images to
  assimilate additional data in order to extraction clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

% Still, one of the major bottlenecks in terms of personalized
% computaional models is the quality of data (and representations of
% model). There is a big issue in
% reproducebility, because of noisy data, and operator depedent
% measurements.
Today there are three main non-invasive imaging modalities used in
cardiology. These are echocardiography (Echo), magnetic
resonance imaging (MRI) and computed tomography (CT). Each
modality offers advantages and disadvantages over the other.
For example, MRI provides
high quality images, uses zero radiation, but is expensive and lacks
temporal resolution. CT can more accurately reconstruct the 3D image
in contrast to MRI, in which 2D slices needs to be glued together to
form a 3D surface. However, CT exposes the patient to radiation, which
increases the chance of developing cancer. Finally echocardiography is
easy to use, cheap, harmless, and  has good temporal resolution, but
is clearly inferior when it comes to image quality.

@@

The main modality used in this thesis is 4D echocardiography, and we
will therefore focus on data acquired using this modality.
With 4D we mean three spatial and one temporal dimension.
The speed of sound in human tissue (which is approximately 1540 m/s)
put some limitations on the image quality versus the frame-rate
\cite{rabben2010technical} for these 3D volume images. In order to
enhance these 3D volume images, disjoint subvolumes acquired during
$N$ cardiac cycles ($N$ typically between 2 and
8)\cite{brekke2007volume} are stitched together.

The images are later processed using some image segmentation tool.
One such tool is EchoPac, which is used for analyzing images acquired with
the echo scanner from GE Vingmed. The 4D Auto LVQ tool in EchoPac, is
a tool for processing 3D echo images, and traces of volume,
triangulated surfaces and 3D strain traces \cite{heimdal20114d},
together with a structured mesh of the American Heart Association (AHA) segments
\cite{cerqueira2002standardized} for each time point can be extracted, see Figure
\ref{fig:echopac_output}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.38\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/bullseye/bullseye.pdf}
    \caption{\label{fig:bullseye_intro}}
  \end{subfigure}
\caption{Figure \ref{fig:echopac_out_surf} shows triangulated surfaces of the
      endo- and epicardium, and Figure
      \ref{fig:echopac_out_strain_mesh} shows structured mesh of the
      American Heart Association (AHA) segments. These surfaces are
      exported from the image segmentation tool EchoPac. Figure 
      \ref{fig:bullseye_intro} shows the American Heart Association (AHA)
      segments illustrated on a so called bullseye plot with name and
      numbers on the different segments. }
\label{fig:echopac_output}
\end{figure}

@@

\subsubsection{Mesh generation}
In this section we will explain how to generate a left ventricular
mesh based on segmented surfaces coming from 4D echocardiography.
Figure \ref{fig:echopac_output} shows an example of how the exported data
from the image segmentation tool looks like. For each time point we
can extract triangulated surfaces of the endocardium and epicardium
which can be seen in Figure \ref{fig:echopac_out_surf}. Together with
these surfaces we are also given a so called strain mesh (Figure
\ref{fig:echopac_out_strain_mesh}) located approximately in the
midwall, which defines the approximate location of the AHA-segments
\cite{cerqueira2002standardized}, and can be used to orient the surfaces.

As discussed in Section \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the positive $x$
direction. To construct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
square fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct cut size is found, the points above the plane is removed,
and the remaining surfaces are smoothed using the mesh generation tool GAMer
\cite{yu2008feature}.

@@

The actual mesh generation is performed using Gmsh
\cite{geuzaine2009gmsh}, which mesh the endocardial and epicardial
surface together using frontal-Delaunay meshing algorithm. Gmsh also marks
the endocardial, the epicardial  and the basal facets, along with the
endocardial and epicardial basal rings.

@@

\begin{remark}
  The cavity volume is computed by generating the whole mesh, and the
  compute the volume using Equation \eqref{eq:volume_operator}. The cut size is
  found by using a one dimensional optimization algorithm with
  the objective functional representing the squared error between
  computed and measured volume.
\end{remark}

\subsubsection{Rule-based fiber architecture}
\label{sec:rule_based_fiber}
Both the passive and active properties of the myocardium depends on
the underlying muscle fiber architecture, and it is therefore of high
relevance to capture this architecture as accurately as
possible. Unfortunately, imaging technology does not currently provide a
way of extracting patients-specific fiber orientation \emph{in
  vivo}. Diffusion Tensor Imaging (DTI) can however be utilized to reconstruct
the fiber and laminar structure \emph{ex vivo}\cite{rohmer2007reconstruction}.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]
  {chapters/introduction/figures/fiber/fiber}
\caption{The LDRB algorithm \cite{bayer2012novel} applied to an LV
  (left) and BiV (right) geometry with an helix fiber angle of
  $60^{\circ}$ and $-60^{\circ}$ on the endo- and epicardium
  respectively. Color represents the helical angle, with blue being
  $0^{\circ}$ (circumferential) which can be seen in the midwall. }
\label{fig:rule_based_fiber}
\end{figure}

The LDRB algorithm takes as
input the geometry $\Omega$, the helical fiber angle on the endo- and
epicardium ($\alpha_{\mathrm{endo}}$ and $\alpha_{\mathrm{epi}}$) and
the transverse fiber angle on the endo- and epicardium
($\beta_{\mathrm{endo}}$ and $\beta_{\mathrm{epi}}$), and outputs a
set of three orthogonal vector fields, one representing the fiber
angle $(\ef)$, one representing the sheet angle $(\eS)$ and the final
direction is called the sheet normal direction $(\en)$. Figure
\ref{fig:rule_based_fiber} shows the output of this algorithm applied
to both an LV and a BiV ellipsoidal geometry.

@@

When dealing with geometric shapes such as a left ventricle, finding a
coordinate system that is easy to work with is important. For
example, when studying spherical shapes, the spherical coordinate system is
usually easier to work with. Likewise, a coordinate system that is
easy to use when studying ellipsoidal shapes like the left ventricle is
the \emph{prolate spheroidal coordinate system} \cite{hunter1996kd}. 
The Cartesian coordinates $(x,y,z)$ are related to the prolate
spheroidal coordinates by

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta  \\
    y &= a \sinh \mu \sin \nu \sin \theta  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuth angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semi-major
axis, and $c$ the semi-minor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. The inverse relation, going from the
Cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$, is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau) \\
    \mu &= \arccosh (\sigma) \\
    \theta &= \arctan \frac{z}{y} \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big) \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big) 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{Illustration of the prolate spheroidal coordinate system.}
\label{fig:prolate}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to obtain the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with some loads $\mathbf{T}$. We want
to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $\mathbf{T}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} with $\mathbf{T}$ representing
the traction on the endocardium, and which depends on the reference
coordinates. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To
find  $\Xvec_{\mathcal{U}}$ we let
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the load $\mathbf{T}$ do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is large. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

@@

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides an a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}.
This algorithm is based on subtracting a factor $k$ times
the displacement obtained after inflating the image-based geometry to
the target pressure. To estimate the
unloaded configuration $\Xvec_{\mathcal{U}}$, we perform the following steps:
\begin{enumerate}
  \item Inflate $\Xvec_{\mathcal{I}}$ to target
    pressure $(p_{\mathcal{I}}^{\mathrm{LV}},
    p_{\mathcal{I}}^{\mathrm{RV}})$ to obtain
    $\mathbf{d} = \mathbf{d}(\Xvec_{\mathcal{I}})$.
  \item Given $k \in \mathbb{R}$, the unloaded configuration is given by
    \begin{align}
      \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
     \end{align}
     Further let $\mathbf{d}_k = \mathbf{d}(\Xvec_{\mathcal{U}}(k))$
     be the displacement obtained after inflating the unloaded
     configuration to the target pressure.
   \end{enumerate}
The uloading algorithm is based on finding $k$ so that the difference
between the image-based geometry and the inflated, unloaded geometry
is minimized, that is 
\begin{align}
  \underset{k}{\text{minimize}}
  \;\;  \mathrm{res} ( \Xvec_{\mathcal{I}}, \Xvec_{\mathcal{U}}(k) + \mathbf{d}_k)
  \label{eq:unloading_Raghavan}
\end{align}
with
\begin{align}
  \mathrm{res}(A, B) = \sup_{a \in A} \inf_{b \in B} d(a, b), 
\end{align}
and $d$ being the Euclidean distance.

% suppose we can write
% \begin{align}
%   \Xvec_{\mathcal{I}} = \Xvec_{\mathcal{U}} + k \cdot \mathbf{d}(\Xvec_{\mathcal{U}}).
% \end{align}
% Of course this assumption might not be true, but we can find a value of
% $k$ that minimizes some residual using a 1D optimization
% algorithm. Note that choosing $k = 1$, would be the same as applying
% one iteration of the backward displacement method.

Since the backward displacement method really does converge towards a
fixed point, it is favorable to the Rahavan method. Therefore, a
hybrid approach is to apply the backward displacement method as
long as the non-linear solver converges, and switch to the Rahavan
method upon divergence.

% For a transversely isotropic material (or orthotropic), the change in
% fiber field when going from one geometry to another, has to be
% considered. The natural thing to do is apply a \emph{Piola
%   transformation} to the fibers in the image based geometry: Let
% $\F_{n}= \I - \Grad \left(\mathbf{d}(\Xvec_n) \right)$ and left
% $\ef^{n} = J^{-1}\F_{n} \ef$ where $\ef$ is the fiber field on
% $\Omega_{\mathcal{I}}$. Another way is to regenerate the fibers (see
% Section \ref{sec:rule_based_fiber}) using the new unloaded geometry as
% input. We will see in the following numerical experiment that non of
% these methods gives the desired result.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stress in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the constitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical systems such as the heart or weather
forecast, it is not always possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

@@

There are basically two main approaches to assimilate data;
\emph{sequential data assimilation} and \emph{variational data
  assimilation}. Sequential data assimilation, which is often
referred to as filtering, is based on a predictor-corrector approach
where you start from your initial data and use a filter to predict the
next state. Once you have your measurement available you correct the
estimate based on the new observations. Some examples of sequential
procedures include Kalman filtering and Luenberger observers
\cite{chapelle2013fundamental}.

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
that is taken in this thesis. For a more thorough review of other
methods we refer to \cite{chapelle2013fundamental}.
